---
title: "Appendix MRes 2021 Assignment (Jan-Philipp Cieslik)"
output:
  pdf_document: 
    keep_tex: yes
  html_notebook: default
  html_document:
    df_print: paged
---
```{r, include=FALSE}
#Libraries for creation of rmarkdown output, not required for actual calculations
library(rmarkdown)
library(knitr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)

#Snippet from https://stackoverflow.com/questions/25646333/code-chunk-font-size-in-rmarkdown-with-knitr-and-latex
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Setup - Bash (Terminal)
Download data sets from Xena Browser (TCGA BRCA) \
https://xenabrowser.net/datapages/?cohort=TCGA%20Breast%20Cancer%20(BRCA)

```{bash, engine.opts='-l', message=FALSE, warning=FALSE, error=FALSE, results='hide', size='small'}
wget -O survival.tsv \
https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/survival%2FBRCA_survival.txt
wget -O clinical_matrix.tsv \
https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.BRCA.sampleMap%2FBRCA_clinicalMatrix
wget -O HiSeqV2.tsv.gz \
https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.BRCA.sampleMap%2FHiSeqV2.gz
wget -O Methylation450k.tsv.gz \
https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.BRCA.sampleMap%2FHumanMethylation450.gz
wget -O Methylation450k_probemap.tsv \
https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/probeMap%2FilluminaMethyl450_hg19_GPL16304_TCGAlegacy
wget -O CNV_thresholded.tsv.gz \
https://tcga-xena-hub.s3.us-east-1.amazonaws.com/download/TCGA.BRCA.sampleMap%2FGistic2_CopyNumber_Gistic2_all_thresholded.by_genes.gz
```


Unzip the downloaded files
```{bash, engine.opts='-l'}
gunzip HiSeqV2.tsv.gz
gunzip CNV_thresholded.tsv.gz
gunzip Methylation450k.tsv.gz
```
# Setup - R
Load libraries
```{r, message=FALSE, warning=FALSE, results='hide'}
library(survival)
#with the given option the fread function from data.table behaves like read.table
#but is much quicker and memory efficient
library(data.table)
options(datatable.fread.datatable=FALSE)
```

# Data Loading - RNA/Survival/Clinical
Load data and adjust row names
```{r}
#set the gene of interest
gene <- "SPIB"
#using check.names=FALSE to prevent the change from hyphens to dots
rna.data <- fread("HiSeqV2.tsv", sep="\t", head=TRUE,
                  stringsAsFactors =FALSE, check.names=FALSE)

rownames(rna.data) <- make.unique(rna.data[, 1])
rna.data <- rna.data[,-1]
rna.data <- as.matrix(rna.data)

surv.data <- read.table("survival.tsv", sep="\t", header=T, row.names=1) 
clin.data <- read.table("clinical_matrix.tsv", sep="\t", header=T, row.names=1, quote = "")
```

Generate survival data
```{r}
os.time <- surv.data[colnames(rna.data),"OS.time"]
os.event <- as.numeric(surv.data[colnames(rna.data),"OS"])
brca.os <- Surv(os.time,os.event)
#Delete local variables (optional, just to keep the environment clean)
rm("os.time", "os.event")
```

# Univariate Regression Analysis (RNA/Survival)
Create empty data frame for results
``` {r}
rna.survival.univariate<-array(NA, c(nrow(rna.data),4))
colnames(rna.survival.univariate)<-c("HR","LCI","UCI","PVAL")
rownames(rna.survival.univariate)<-rownames(rna.data)
rna.survival.univariate<-as.data.frame(rna.survival.univariate)
```

Iterate through all genes and generate Cox model
```{r, warning=FALSE, message=FALSE}
for(i in 1:nrow(rna.data))
{
  #Check if less than 2 samples are available for correlation
  if(sum(!is.na(rna.data[i,])) < 2){
    next
  }
 coxphmodel <- coxph(brca.os ~ as.numeric(rna.data[i,]))
 summary <-summary(coxphmodel)
 rna.survival.univariate$HR[i] <- summary$coef[1,2]
 rna.survival.univariate$LCI[i] <- summary$conf.int[1,3]
 rna.survival.univariate$UCI[i] <- summary$conf.int[1,4]
 rna.survival.univariate$PVAL[i] <- summary$coef[1,5]
}
rna.survival.univariate <- as.data.frame(rna.survival.univariate)
rna.survival.univariate$FDR <- p.adjust(rna.survival.univariate$PVAL,method="fdr")
rna.survival.univariate <- 
  rna.survival.univariate[order(rna.survival.univariate$FDR, decreasing=F),]

#Remove local variables
rm("summary", "i")
```
Print results of univariate analysis
```{r, results='asis'}
kable(rna.survival.univariate[1:5,])
```
# Multivariate Regression Analysis (RNA/Survival/Clinical)
Clinical data preparation
```{r}
#subset clinical data to patients that also have RNA data
clin.data<-clin.data[colnames(rna.data),]
#create age variable
age<-as.numeric(clin.data$age_at_initial_pathologic_diagnosis)
#create stage high/low variable
x3<-grep("III",clin.data$Converted_Stage_nature2012)
x4<-grep("IV",clin.data$Converted_Stage_nature2012)
stage.high<-rep(0,nrow(clin.data))
stage.high[c(x3,x4)]<-1

#Remove local variable
rm("x3", "x4")
```

Create empty data frame for results
```{r}
rna.survival.multivariate<-array(NA, c(nrow(rna.data),4))
colnames(rna.survival.multivariate)<-c("HR","LCI","UCI","PVAL")
rownames(rna.survival.multivariate)<-rownames(rna.data)
rna.survival.multivariate<-as.data.frame(rna.survival.multivariate)
```

Iterate through all genes to generate multivariate regression model
```{r, warning=FALSE, message=FALSE}
for(i in 1:nrow(rna.data))
{
  #Check if less than 2 samples are available for correlation
  if(sum(!is.na(rna.data[i,])) < 2){
    next
  }
 coxphmodel <- coxph(brca.os ~ rna.data[i,]+age+stage.high)
 summary <- summary(coxphmodel)
 rna.survival.multivariate$HR[i] <- summary$coef[1,2]
 rna.survival.multivariate$LCI[i] <- summary$conf.int[1,3]
 rna.survival.multivariate$UCI[i] <- summary$conf.int[1,4]
 rna.survival.multivariate$PVAL[i] <- summary$coef[1,5]
}
rna.survival.multivariate <- as.data.frame(rna.survival.multivariate)
rna.survival.multivariate$FDR <- p.adjust(rna.survival.multivariate$PVAL,method="fdr")
rna.survival.multivariate <-
  rna.survival.multivariate[order(rna.survival.multivariate$FDR, decreasing=F),]
#Remove local variables
rm("summary", "i")
```

Print results of multivariate analysis
```{r, results='asis'}
kable(rna.survival.multivariate[1:5,])
```

```{r}
gene.info <- rna.survival.multivariate[gene,]
gene.high <- as.numeric(rna.data[gene,]>median(rna.data[gene,]))

#calculate log rank test
gene.high.logrank <- survdiff(brca.os ~ gene.high)
gene.high.logrank.p <- 1 - pchisq(gene.high.logrank$chisq, length(gene.high.logrank$n) - 1)
print(survfit(brca.os ~ gene.high))

#create survival plot
plot.text <- paste0("HR = ",round(gene.info["HR"], digits = 2),
                    " (q = ",round(gene.info["FDR"], digits = 5),")")
plot.text2 <- paste0("log-rank: p = ",round(gene.high.logrank.p, digits = 2))
plot.legend <- c(paste0(gene,"-high"),paste0(gene,"-low"))


plot(survfit(brca.os ~ gene.high), col=c("black","red"), lwd=2, 
     mark.time=TRUE, xlab="Time (Years)", ylab="Overall Survival",
     xscale = 365.25, xmax = 15*365.25)
legend("topright",legend=plot.legend,col=c("red","black"),lwd=2)
text(0,0.1,plot.text, adj = c(0,0))
text(0,0,plot.text2, adj = c(0,0))

#delete local variables
rm("plot.text", "plot.text2", "plot.legend")
```

# Methylation data
Load methylation data
```{r}
meth.annotation <- read.table("Methylation450k_probemap.tsv", sep="\t",
                              header=T, comment.char="")
meth.data <- fread("Methylation450k.tsv", sep="\t", header=TRUE)

row.names(meth.data) <- meth.data[,1]
row.names(meth.annotation) <- meth.annotation[,1]
meth.data <- meth.data[, -1]
meth.annotation <- meth.annotation[, -1]
meth.probes<- rownames(meth.annotation[grep(gene, meth.annotation$gene),])
```
Display methylation data
```{r, results='asis'}
kable(meth.data[1:3,1:3])
kable(meth.annotation[1:3,])
```

subset data set to samples that have rna, methylation and survival data.
```{r}
samples <- intersect(colnames(meth.data),colnames(rna.data))
meth.intersect.rna <- meth.data[,samples]
rna.intersect.meth <- rna.data[,samples]
surv.intersect.rna.meth <-
  surv.data[intersect(rownames(surv.data),colnames(rna.intersect.meth)),]

meth.intersect.rna <- as.matrix(meth.intersect.rna)
rna.intersect.meth <- as.matrix(rna.intersect.meth)
surv.intersect.rna.meth <-
  as.data.frame(surv.intersect.rna.meth[colnames(meth.intersect.rna),])

meth.intersect.rna <-  meth.intersect.rna[meth.probes,,drop=FALSE]

rm("samples")
```

```{r}
#exclude methylation sites that are not determined in more than 0.5 of samples
na.count <- apply(meth.intersect.rna,1,function(x) sum(as.numeric(is.na(x))))
exclude <- as.numeric(na.count>0.5*ncol(meth.intersect.rna))
meth.intersect.rna <- meth.intersect.rna[which(exclude==0),, drop=FALSE]

#generate empty array for results
results.meth<-array(NA,c(nrow(meth.intersect.rna),5))
rownames(results.meth)<-rownames(meth.intersect.rna)
colnames(results.meth)<-c("Cor","pval","qval","Mean.high","Mean.low")
gene.high.meth.rna <- as.numeric(
  as.numeric(rna.intersect.meth[gene,]) > median( as.numeric(rna.intersect.meth[gene,]) )
  )

#remove local variables
rm("na.count", "exclude")
```

Iterate through every methylation site of the selected gene and perform correlation.
```{r}
for (i in 1:nrow(meth.intersect.rna))
{
  results.meth[i,1] <-
    cor.test(as.numeric(rna.intersect.meth[gene,]),as.numeric(meth.intersect.rna[i,]),
             use="c", method = "spearman",exact=FALSE)$est
  results.meth[i,2] <-
    cor.test(as.numeric(rna.intersect.meth[gene,]),as.numeric(meth.intersect.rna[i,]),
             use="c", method = "spearman",exact=FALSE)$p.value
}
results.meth[,4] <- apply(meth.intersect.rna[,which(gene.high.meth.rna==1), drop=FALSE],
                          1,mean,na.rm=T)
results.meth[,5] <- apply(meth.intersect.rna[,which(gene.high.meth.rna==0), drop=FALSE],
                          1,mean,na.rm=T)
results.meth[,3] <- p.adjust(results.meth[,2],method="fdr")
results.meth<-results.meth[order(results.meth[,3], decreasing=F),,drop=FALSE]

#remove local variables
rm("i")
```

```{r, include=FALSE}
for(i in 1:5){
  results.meth[,i] <- signif(results.meth[,i], digits=3)
}
write.table(results.meth, file="results.meth.csv", eol = "\\\\ \n", sep = " & ", quote=FALSE)
```

Display results
```{r, results='asis'}
kable(results.meth)
```

```{r}
#set the methylation site of interest
meth.site = "cg07979271"
plot.title <- paste0(gene, " in BRCA")
plot.text <- paste0("cor = ",round(results.meth[meth.site, "Cor"], digits = 2),
                    " (q = ",format(results.meth[meth.site, "qval"], scientific = TRUE),")")
plot.legend <- c(paste0(gene,"-high"),paste0(gene,"-low"))

 
plot(as.numeric(meth.intersect.rna[meth.site,]), as.numeric(rna.intersect.meth[gene,]), 
     xlab=paste0("Methylation Level (", meth.site,")"), ylab="mRNA Level")
text(0.45,0,plot.text, adj = c(0,0.5))
abline(lm(rna.intersect.meth[gene,] ~ meth.intersect.rna[meth.site,]))

```

# CNV data
Load CNV data and display table
```{r, results='asis'}
cnv.data <- fread("CNV_thresholded.tsv", sep="\t", header=T)
rownames(cnv.data) <- make.unique(cnv.data[,1])
cnv.data <- cnv.data[,-1]
cnv.data <- as.matrix(cnv.data)
#subset data set to entries that are also available in the RNA data set
cols.intersect <- intersect(colnames(cnv.data), colnames(rna.data))
row.intersect <- intersect(rownames(cnv.data), rownames(rna.data))
cnv.intersect.rna <- cnv.data[row.intersect, cols.intersect]
rna.intersect.cnv <- rna.data[row.intersect, cols.intersect]
kable(cnv.data[1:5,1:3])
rm("cols.intersect","row.intersect")
```
Calculate CNV changes in selected gene
```{r}
cnv.rna.df <- data.frame(CNV = cnv.intersect.rna[gene,],
                         RNA = rna.intersect.cnv[gene,],
                         stringsAsFactors=FALSE)
cnv.rna.df[which(cnv.rna.df$CNV == -2), "CNV"] <- "deep loss"
cnv.rna.df[which(cnv.rna.df$CNV == -1), "CNV"] <- "shallow loss"
cnv.rna.df[which(cnv.rna.df$CNV == 0), "CNV"] <- "diploid"
cnv.rna.df[which(cnv.rna.df$CNV == 1), "CNV"] <- "low-level gain"
cnv.rna.df[which(cnv.rna.df$CNV == 2), "CNV"] <- "high-level gain"
#"deep loss" is missing on purpose,
#as no entry falls into this category for our candidate gene
cnv.rna.df$CNV <- factor(cnv.rna.df$CNV,
                         levels=c("shallow loss", "diploid",
                                  "low-level gain", "high-level gain"),
                         ordered=TRUE)

aov(RNA ~ CNV, data = cnv.rna.df)

boxplot(RNA~CNV, data=cnv.rna.df,
   xlab="Copy Number", ylab="mRNA Level")
```

# mRNA correlation
Create empty data frame for results
```{r}
rna.cor <-array(NA,c(nrow(rna.data),5))
rownames(rna.cor)<-rownames(rna.data)
colnames(rna.cor)<-c("Cor","pval","qval","Mean.high","Mean.low")

gene.high.meth.rna <- as.numeric(
  as.numeric(rna.data[gene,]) > median( as.numeric(rna.data[gene,]) )
  )
```
Loop through every gene and correlate it with the candidate gene
```{r}

for (i in 1:nrow(rna.data)){
  #Check if less than 2 samples are available for correlation
  if(sum(!is.na(rna.data[i,]) & !is.na(rna.data[gene,])) < 2){
    next
  }
  #Check if all values are zero
  if(sum(rna.data[i,]) == 0){
    next
  }
  if(sum(!is.na(rna.data[i,])))
  result.temp <-
    cor.test(as.numeric(rna.data[gene,]), as.numeric(rna.data[i,]),
             use="c", method = "spearman", exact=FALSE)
  rna.cor[i, 1:2] <- c(result.temp$est, result.temp$p.value)
}
rna.cor[,4] <- apply(rna.data[,which(gene.high.meth.rna==1), drop=FALSE],1,mean,na.rm=T)
rna.cor[,5] <- apply(rna.data[,which(gene.high.meth.rna==0), drop=FALSE],1,mean,na.rm=T)
rna.cor[,3] <- p.adjust(rna.cor[,2],method="fdr")
rna.cor <- rna.cor[order(rna.cor[,3], decreasing=F),,drop=FALSE]
```
Display results
```{r, results='asis'}
kable(rna.cor[1:5,])
```

```{r, include=FALSE}
#save results (for LaTeX table generation)
for(i in 1:5){
  rna.cor[,i] <- signif(rna.cor[,i], digits=3)
}
rna.cor2 <- rna.cor[order(rna.cor[,1], decreasing=F),,drop=FALSE]
write.table(rna.cor2, file="results.rna2.csv", eol = "\\\\ \n", sep = " & ", quote=FALSE)
rm("rna.cor2")
```

Save workspace 
```{r}
save.image("assignment.RData")
```
